\PaperTitle

### 2020-01-19 paper/filter_bubbles.tex
updates paper

\documentclass[sigconf]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage{float}
\usepackage{enumitem}
\newtheorem{finding}{Finding}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2020}
\acmYear{2020}
\setcopyright{acmlicensed}
\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipei, Taiwan}
\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmDOI{10.1145/1122445.1122456}
\acmISBN{978-1-4503-9999-9/18/06}

%
% These commands are for a JOURNAL article.
%\setcopyright{acmcopyright}
%\acmJournal{TOG}
%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
%\acmDOI{10.1145/1122445.1122456}

%
% Submission ID. 
% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
% of the event, and this ID should be used as the parameter to this command.
%\acmSubmissionID{123-A56-BU3}

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
%\citestyle{acmauthoryear}

%
% end of the preamble, start of the body of the document source.
\begin{document}
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{0pt}
% end of the preamble, start of the body of the document source.

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.

\settopmatter{printacmref=false} % Removes citation information below abstract
\makeatletter
\def\-copyrightspace{\relax}
\makeatother

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
We study a model of user decision-making in the context of recommender systems. We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems. The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products. We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al. (2014). However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
\par
We discuss the evaluation of recommender systems in light of these findings. Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved and ignore user beliefs and the inferences that users themselves make after consuming items. We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided. As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%
\author{Guy Aridor}
\affiliation{
  \institution{Columbia University \\ Department of Economics}}

\author{Duarte Gon\c{c}alves}
\affiliation{%
  \institution{Columbia University, BRU-IL \\ Department of Economics}}

  \author{Shan Sikdar}
  \affiliation{Independent Researcher}
%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
%\keywords{Recommender Systems, Filter Bubbles, Serendipity}

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.

@@

\section{Introduction}

Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets. For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption. Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
\par

Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar. 12), YouTube Extremism and the Long Tail, \textit{The Atlantic}. \url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par

\noindent \textbf{Our Model and Contributions} In this paper, 
%drawing from recent work in psychology and decision theory in economics, 
we first propose a model of consumer decision-making in the context of large choice sets faced by users in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations. 
We then ask how a stylized model of recommendation can shape the behavior of long-lived users
and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
\par

The key components of our model are:
\begin{enumerate}[label=(\arabic*)]
\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under uncertainty.
\item Choice sets are large and users only consume a small fraction of this choice set over their lifetime.
\item Consumption of these items is sequential and users live for multiple periods.
\item Consuming an item provides information about the value of similar items in the product space.
\item Recommendation provides users with information, reducing their uncertainty.
\end{enumerate}

Using this model we first investigate the degree to which recommender systems lead users into filter bubbles. Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}. The driving force of this in our model is that preferences for similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty. Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space. This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
We compare this benchmark to the case where recommendations are provided. 
We consider recommendations as giving information to users in order to reduce their uncertainty.
Recommendations could potentially make matters worse: providing information about the value of the goods when this value is correlated could intensify local search and decrease consumption diversity. However, and in line with the empirical results from \cite{nguyen2014exploring}, we find that recommendations reduce the narrowing effect.
\par

We then explore how diverse the overall set of items that users consume is. It could be expected that higher diversity being associated with more intense search is associated with higher welfare. Much to the opposite, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of values, leading to higher overall value from consumption. When they consume products they value little, they move to another part of the product space. A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity. Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users. 
\par

We find that recommendation can lead to increases in homogeneity among users. We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component. This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data. The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation. This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
\par

Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems. Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation. Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior. If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation. Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space. Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in the face of risk-aversion.
\par

\textbf{Related Work.} 
\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content. A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}. The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption. They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume. To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies. Our paper provides a theoretical framework through which to view this problem.
The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other. \cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems. Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here. \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g. \cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations. Our evaluation measure is most closely related to the notion of serendipity in recommender systems that has been studied extensively (e.g. see \cite{kotkov2016survey} for a survey).

\section{Our Model and Preliminaries}

\subsection{Preliminaries on Expected Utility Theory}
\noindent For every product $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each consumer $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it. Each consumer can value the same product in differently. However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous. So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it. In particular, even users that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it. We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space. Note that this implies that consuming product $n$ is the same as taking a gamble. Each individual evaluates the product according to the associated expected utility associated with the product, i.e. $U_i(n)=\mathbb E_{p_i}[u(x_n)]$. 
\par

Risk aversion captures the how different individuals react to the risk associated to a particular consumption opportunity. It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the individual indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$. An individual $i$ is more risk-averse than another individual $j$ when for every gamble $x$, individual $i$ requires a lower value to give up the gamble when compared to individual $j$, i.e. $\delta_i(x)$. Therefore, a more risk-averse individual is more willing to avoid the risk of taking the gamble. We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences. Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
\par

@@

We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$. In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
\par
Stacking values in vector-form, we get the vector of values associated with each item 
$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
\par
\noindent\textbf{User Decision-Making.}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off. However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
\par
\noindent \textbf{User Beliefs.} We assume that users do not know the realized utilities of the goods before consuming an item.
Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal 
\begin{enumerate}[topsep=0pt]
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}
We impose the normality assumption for two reasons. The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed. These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over good $n$. As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
\par

We assume users are expected utility maximizers. User $i$'s certainty equivalent for product $n$, the sure value that makes user $i$ indifferent between it and consuming the product $n$, conditional on the consumption history, is given by
$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.

\noindent \textbf{User Learning}
When a consumer consumes a product $n$ they learn the realized value for that product. In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items. 
%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
We assume that learning about the value of good $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
\par

>>>>

\documentclass[format=acmsmall, review=false]{acmart}
\usepackage{acm-ec-20}
\usepackage{booktabs} % For formal tables
\usepackage[ruled]{algorithm2e} % For algorithms

\usepackage{subcaption}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}

\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Choose a citation style by commenting/uncommenting the appropriate line:
%\setcitestyle{authoryear}
\setcitestyle{acmnumeric}

\newtheorem{finding}{Finding}

% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
\newcommand\PaperTitle{Consumer Decision-Making and Recommender Systems}
\title[\PaperTitle]{\PaperTitle}

% Anonymized submission.
\author{Submission 42?}

% Abstract. Note that this must come before \maketitle.
\begin{abstract}
We study a model of user decision-making in the context of recommender systems. We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems. The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products. We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al. (2014). However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
\par
We discuss the evaluation of recommender systems in light of these findings. Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved and ignore user beliefs and the inferences that users themselves make after consuming items. We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided. As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
\end{abstract}

\begin{document}

@@

% Paper body
\section{Introduction}

Users in many online markets face difficult choice problems as they have to sort through the vast choice sets present in these markets. For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. A common remedy by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions by providing users with information and recommendations to allow them to make more informed consumption decisions. These systems have been shown to be influential in steering consumer choice with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. 
\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. 
Retrieved on October 3, 2019.}
\par 
Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices. Understanding this has become increasingly important as there have emerged a number of concerns regarding the consequences of such systems. For instance, there have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar. 12), YouTube Extremism and the Long Tail, 
\textit{The Atlantic}. 
\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} 
that personalized recommender systems lead users into \textit{filter bubbles} 
where they effectively get isolated from a diversity of viewpoints or content 
\cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}. Further, an emerging worry is that recommender systems can be utilized as a tool of market power by steering consumer choice towards products more profitable for online platforms to the detriment of consumer welfare [citations]. Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}. 
\par
While these issues are important, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology. It is difficult to understand to what extent issues such as filter bubbles and homogenization are a consequence of recommender systems or a natural consequence of consumer decision-making in these contexts without such analysis. Further, it is challenging to understand how recommender systems can be used anti-competitively without understanding how such steering can influence user behavior. Finally, a major open question in the literature on incentivizing exploration in recommender systems is how to design recommendation policies in the context of realistic user behavior \cite{immorlica2018incentivizing}, which requires an understanding of what models of user decision-making are appropriate.

\xhdr{Our Model} In this paper we study a model of consumer decision-making in these markets via means of simulation and ask how a stylized model of recommendation can shape the behavior of users.
\par
The first component of our model is that users sequentially consume items and face large choice sets, though they only consume a small fraction of this choice set over their lifetime.
\par
The second component of our model is that users have \textit{uncertainty} about their valuation of the different items. This is motivated both by the fact that many markets where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place. Thus, users face a sequential decision-making problem under uncertainty.
\par 
The third, and key, component of our model is that users are Bayesian and employ similarity-based generalization to navigate the space of items. Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved. This valuation provides information for consumers that enables them to update their beliefs about similar items. This exploits the fact that similar items have more correlated true valuations, which enables users to navigate the vast product space.
\par 
The fourth component of our model is that recommendation provides users with uncertainty about the true valuations. We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component. This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommender can learn from previous users' data. The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.

\xhdr{Main Findings}
Our first set of results focuses on understanding to what extent recommender systems lead users into filter bubbles and how such filter bubbles may naturally arise. The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time. However, empirical work, such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects without recommendation and that, in fact, recommender systems mitigate such effects. The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
\par 
We find that, even if there were no uncertainty about valuations, if there is no inherent correlation between the true valuations of the items then such filter bubble effects never arise, even in the optimal consumption path. However, once there is a correlation between the true valuations of the goods then such filter bubble effects arise even without recommendation.\footnote{The correlation between the true valuation of items in the product space is a natural assumption and, indeed, is the core idea behind standard content-based recommendation methods. These methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about the set of items and when they begin consuming items this impacts the beliefs of the average utility of nearby items. When a user finds an item that has high realized utility, this induces them to update their beliefs positively about the realized utility of nearby items which leads to them consuming these items. Thus, without recommendation users increasingly consume items in this portion of the vast product space since they never get information about items in different parts of the product space. This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items. Thus, this effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
\par 
We then investigate whether the introduction of recommendation either mitigates or amplifies this effect. Recommendations could potentially make matters worse: providing information about the common value of the items when this value is correlated could intensify the extent of local consumption. However, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
\par
\xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume. Consumption diversity is increasingly a common goal of many recommender systems [references], which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare. However, we find that, without recommendation, higher diversity of items does not ensure higher welfare. This is due to the fact that high consumption set diversity in our model can arise from a sequence of bad realizations, which leads to very intense search over very different parts of the product space and high consumption diversity but very low welfare. However, discovering a sweet spot leads to highly rewarding local search and very low diversity. Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on user welfare.
\par
We then turn to the extent to which recommendation leads to increasing user homogenization. We find that, without recommendation, there is very little homogenization as users are effectively all doing their own random exploration around the product space. In the ex-post utility maximizing consumption path (i.e. the consumption path the user would have taken if all uncertainty had been resolved) we find that some degree of homogenization is optimal. However, we find that while recommendation leads to an increase in consumer welfare, it concentrates user consumption choices in the same portions of the product space where there is a high common-value component. As a result, it leads to a substantially higher homogenization when compared both to the no recommendation and ex-post utility maximizing consumption path.
\par 
One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow. This is important for the debate regarding the extent to which recommender systems can be utilized as tools of market power since most of the literature analyzes this in a static context. However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.

\xhdr{Related Work.} 
\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content. A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}. The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption. They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume. To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies. Our paper provides a theoretical framework through which to view this problem.
The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other. \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems. Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here. \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.

\section{Our Model and Preliminaries}
\subsection{Preliminaries on Expected Utility Theory}
\noindent For every product $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each consumer $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it. Each consumer can value the same product in differently. However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous. So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it. In particular, even users that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it. We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space. Note that this implies that consuming product $n$ is the same as taking a gamble. Each individual evaluates the product according to the associated expected utility associated with the product, i.e. $U_i(n)=\mathbb E_{p_i}[u(x_n)]$. 
\par

Risk aversion captures the how different individuals react to the risk associated to a particular consumption opportunity. It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the individual indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$. An individual $i$ is more risk-averse than another individual $j$ when for every gamble $x$, individual $i$ requires a lower value to give up the gamble when compared to individual $j$, i.e. $\delta_i(x)$. Therefore, a more risk-averse individual is more willing to avoid the risk of taking the gamble. We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences. Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one. Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
\par

@@

We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$. In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
\par
Stacking values in vector-form, we get the vector of values associated with each item 
$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
\par
\xhdr{User Decision-Making.}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off. However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
\par
\xhdr{User Beliefs.} We assume that all the true realized utilities are drawn at $t = 0$. However, users do not know the realized utilities before consuming an item, but rather have beliefs over the realized utilities.
Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal 
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}
We impose the normality assumption for two reasons. The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed. These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over good $n$. As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
\par

We assume users are expected utility maximizers. User $i$'s certainty equivalent for product $n$, the sure value that makes user $i$ indifferent between it and consuming the product $n$, conditional on the consumption history, is given by
$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$. Note that this expression is known to be the certainty equivalent for CARA preferences in a Gaussian environment \cite{mas1995microeconomic}.

\xhdr{User Learning}
When a user consumes a product $n$ they learn the realized value for that product. In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items. 
%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
We assume that learning about the value of good $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
\par
